Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-04-08 talk/icooolps2009/paper.tex
- finally insert the Python benchmark  - kill parts of Section 4 and move other parts to the future work section, as    recommended by Michael  - some more fixes and a new reference

\textfloatsep 12pt plus 2pt minus 4pt

@@

One of the often-cited drawbacks of dynamic languages is the performance
penalties they impose. Typically they are slower than statically typed
languages. Even though there has been a lot of research into improving the
performance of dynamic languages (in the SELF project, to name just one example
\cite{XXX}), those techniques are not as widely
used as one would expect. Many dynamic language implementations use completely
straightforward bytecode-interpreters without any advanced implementation
techniques like just-in-time compilation. There are a number of reasons for
this. Most of them boil down to the inherent complexities of using compilation.
Interpreters are simple to implement, understand, extend and port whereas writing a
just-in-time compiler is an error-prone task that is made even harder by the
dynamic features of a language.

@@

The contributions of this paper are:
\begin{zitemize}
\item Applying a tracing JIT compiler to an interpreter.
\item Finding techniques for improving the generated code.
\end{zitemize}

@@

Tracing JITs are built on the following basic assumptions:
\begin{zitemize}
 \item programs spend most of their runtime in loops
 \item several iterations of the same loop are likely to take similar code paths
\end{zitemize}

@@

Typically, programs executed by a tracing VMs goes through various phases:
\begin{zitemize}
\item Interpretation/profiling
\item Tracing
\item Code generation
\item Execution of the generated code
\end{zitemize}

@@

\subsection{Various Issues}

This section will look at some other implementation issues and optimizations
that we have done that are beyond the scope of this paper (and will be the subject
of a later publication).

\textbf{Assembler Backends:} The tracing interpreter uses a well-defined
interface to an assembler backend for code generation. This makes it possible to
easily port the tracing JIT to various architectures (including, we hope, to
virtual machines such as the JVM where backend could generate bytecode at
runtime). At the moment the only implemented backend is a 32-bit
Intel-x86 backend.

\textbf{Trace Trees:} This paper ignored the problem of guards that fail often
because there are several equally likely paths through
a loop. Always falling back to interpretation in this case is not practicable.
Therefore, if we find a guard that fails often enough, we start tracing from
there and produce efficient machine code for that case, instead of always
falling back to interpretation.

\textbf{Allocation Removal:} A key optimization for making the approach
produce good code for more complex dynamic language is to perform escape
analysis on the loop operation after tracing has been performed. In this way all
objects that are allocated during the loop and do not actually escape the loop do
not need to be allocated on the heap at all but can be exploded into their
respective fields.  This is very helpful for dynamic languages where primitive
types are often boxed, as the constant allocation of intermediate results is
very costly.

@@

In this section we try to evaluate the work done so far by looking at some
benchmark numbers. Since the work is not finished, these benchmarks can only be
preliminary. Benchmarking was done on an otherwise idle machine with a 1.4
GHz Pentium M processor and 1 GB RAM, using Linux 2.6.27. All benchmarks where
run 50 times, each in a newly started process. The first run was ignored. The
final numbers were reached by computing the average of all other runs, the
confidence intervals were computed using a 95\% confidence level. All times
include the running of the tracer and machine code production to measure how
costly those are.

The first round of benchmarks (Figure \ref{fig:bench1}) are timings of the
example interpreter (Figure \ref{fig:tlr-basic}) used in this paper computing
the square of 46340 (the largest number whose square still fits into a 32
bit word) using the bytecode of Figure \ref{fig:square}. The results for various
constellations are as follows:

@@

\textbf{Benchmark 2:} The tracing JIT is enabled, but no interpreter-specific
hints are applied. This corresponds to the trace in Figure
\ref{fig:trace-normal}.  The threshold when to consider a loop to be hot is 40
iterations.  As expected, this is not faster than the previous number. It is
even quite a bit slower, probably due to the overheads of the JIT, as well as
non-optimal generated machine code.

@@

\textbf{Benchmark 5:} Same as before, but with the threshold set so high that the tracer is
never invoked to measure the overhead of the profiling. For this interpreter
it seems to be rather large, with 50\% slowdown due to profiling. This is because the interpreter 
is small and the opcodes simple. For larger interpreters (e.g. the Python one) it seems 
likely that the overhead is less significant.

\textbf{Benchmark 6:} Runs the whole computation on the tracing interpreter for estimating the
involved overheads of tracing. The trace is not actually recorded (which would be a
memory problem), so in reality the number is even higher. Due to the double
interpretation, the overhead is huge. It remains to be seen whether that will be
a problem for practical interpreters.

\textbf{Benchmark 7:} For comparison, the time of running the interpreter on top of CPython
(version 2.5.2).

\begin{figure}
\noindent
{\small
\begin{tabular}{|l|r|}
\hline
 &ratio\tabularnewline
\hline
Interpreter compiled to C, no JIT &1\tabularnewline \hline
Normal Trace Compilation &1.20\tabularnewline \hline
Unfolding of Language Interpreter Loop &XXX\tabularnewline \hline
Full Optimizations &0.17\tabularnewline \hline
Profile Overhead &1.51\tabularnewline \hline
Interpreter run by Tracing Interpreter &860.20\tabularnewline \hline
Interpreter run by CPython &256.17\tabularnewline \hline
\end{tabular}
}
\label{fig:bench1}
\caption{Benchmark results of example interpreter computing the square of
46340}
\end{figure}

XXX insert some Python benchmarks

@@

Classical partial evaluation has failed to be useful for dynamic language for
much the same reasons why ahead-of-time compilers cannot compile them to
efficient code. If the partial evaluator knows only the program it simply does
not have enough information to produce good code. Therefore some work has been
done to do partial evaluation at runtime. One of the earliest works on runtime
specialisation is Tempo for C \cite{consel_general_1996, consel_uniform_1996}.
However, it is essentially a normal
partial evaluator ``packaged as a library''; decisions about what can be
specialised and how, are pre-determined. Another work in this direction is DyC
\cite{grant_dyc:expressive_2000}, another runtime specializer for C. Both of these projects
have a problem similar to that of DynamoRIO.  Targeting the C language makes
higher-level specialisation difficult (e.g.\ \texttt{malloc} cannot be
optimized).

@@

Of course there is a lot of work still left to do XXX fix this phrase. Various optimizations are not
quite finished. Both tracing and leaving machine code are very slow due to a
double interpretation overhead and we might need techniques for improving those.
Furthermore we need to apply the JIT to the various interpreters that are
written in RPython to evaluate how widely applicable the described techniques
are. Possible targets for such an evaluation would be the SPy-VM, a Smalltalk
implementation \cite{bolz_back_2008}, a Prolog interpreter or PyGirl, a Gameboy
emulator \cite{XXX}; but also not immediately obvious ones, like Python's
regular expression engine.

If these experiments are successful we hope that we can reach a point where it
becomes unnecessary to write a language specific JIT compiler and just apply a
couple of hints to the interpreter to get reasonably good performance with
relatively little effort.

>>>>

@@

One of the often-cited drawbacks of dynamic languages is the performance
penalties they impose. Typically they are slower than statically typed
languages. Even though there has been a lot of research into improving the
performance of dynamic languages (in the SELF project, to name just one example
\cite{hlzle_adaptive_1994}), those techniques are not as widely
used as one would expect. Many dynamic language implementations use completely
straightforward bytecode-interpreters without any advanced implementation
techniques like just-in-time compilation. There are a number of reasons for
this. Most of them boil down to the inherent complexities of using compilation.
Interpreters are simple to implement, understand, extend and port whereas writing a
just-in-time compiler is an error-prone task that is made even harder by the
dynamic features of a language.

@@

The contributions of this paper are:
\begin{itemize}
\item Applying a tracing JIT compiler to an interpreter.
\item Finding techniques for improving the generated code.
\end{itemize}

@@

Tracing JITs are built on the following basic assumptions:
\begin{itemize}
 \item programs spend most of their runtime in loops
 \item several iterations of the same loop are likely to take similar code paths
\end{itemize}

@@

Typically, programs executed by a tracing VMs goes through various phases:
\begin{itemize}
\item Interpretation/profiling
\item Tracing
\item Code generation
\item Execution of the generated code
\end{itemize}

@@

Machine code production is done via a well-defined interface to an assembler
backend. This makes it possible to easily port the tracing JIT to various
architectures (including, we hope, to virtual machines such as the JVM where
backend could generate bytecode at runtime). At the moment the only implemented
backend is a 32-bit Intel-x86 backend.

@@

In this section we try to evaluate the work done so far by looking at some
benchmark numbers. Since the work is not finished, these benchmarks can only be
preliminary. Benchmarking was done on an otherwise idle machine with a 1.4
GHz Pentium M processor and 1 GB RAM, using Linux 2.6.27. All benchmarks where
run 50 times, each in a newly started process. The first run was ignored. The
final numbers were reached by computing the average of all other runs, the
confidence intervals were computed using a 95\% confidence level. All times
include the running of the tracer and machine code production.

The first round of benchmarks (Figure \ref{fig:bench1}) are timings of the
example interpreter (Figure \ref{fig:tlr-basic}) used in this paper computing
the square of 10000000 (the result will overflow, but for smaller numbers the
running time is not long enough to sensibly measure it)
using the bytecode of Figure \ref{fig:square}. The results for various
constellations are as follows:

@@

\textbf{Benchmark 2:} The tracing JIT is enabled, but no inter\-preter-specific
hints are applied. This corresponds to the trace in Figure
\ref{fig:trace-normal}.  The threshold when to consider a loop to be hot is 40
iterations.  As expected, this is not faster than the previous number. It is
even quite a bit slower, probably due to the overheads of the JIT, as well as
non-optimal generated machine code.

@@

\textbf{Benchmark 5:} Same as before, but with the threshold set so high that the tracer is
never invoked. In this way the overhead of the profiling is measured. For this interpreter
it seems to be rather large, with about 20\% slowdown due to profiling. This is because the interpreter 
is small and the opcodes simple. For larger interpreters (e.g. the Python one) it seems 
likely that the overhead is less significant.

\begin{figure}
\noindent
\begin{tabular}{|l|l|r|r|}
\hline
& &Time (ms) &speedup\\
\hline
1 &compiled to C, no JIT &442.7 $\pm$ 3.4 &1.00\\
2 &Normal Trace Compilation &1518.7 $\pm$ 7.2 &0.29\\
3 &Unrolling of Interp. Loop &737.6 $\pm$ 7.9 &0.60\\
4 &JIT, Full Optimizations &156.2 $\pm$ 3.8 &2.83\\
5 &Profile Overhead &515.0 $\pm$ 7.2 &0.86\\
\hline
\end{tabular}
\label{fig:bench1}
\caption{Benchmark results of example interpreter computing the square of
10000000}
\end{figure}

Furthermore, to test the technique on a more realistic example, we did some
preliminary benchmarks with PyPy's Python interpreter. The function we
benchmarked as well as the results can be seen in Figure
\ref{fig:bench-example}. The function is a bit arbitrary but executing it is
still non-trivial, as a normal Python interpreter needs to dynamically dispatch
nearly all of the involved operations (like indexing into the tuple, addition
and comparison of \texttt{i}). We benchmarked PyPy's Python interpreter with the
JIT disabled, with the JIT enabled and
CPython\footnote{\texttt{http://python.org}} 2.5.4 (the reference implementation of
Python).

The results show that the tracing JIT speeds up the execution of this Python
function significantly, even outperforming CPython by a bit. The tracer needs to
trace through quite a bit of dispatching machinery of the Python interpreter to
achieve this, XXX.

\begin{figure}
\label{fig:bench-example}
{\small
\begin{verbatim}
def f(a):
    t = (1, 2, 3)
    i = 0
    while i < a:
        t = (t[1], t[2], t[0])
        i += t[0]
    return i
\end{verbatim}
}
\noindent
\begin{tabular}{|l|l|r|r|}
\hline
& &Time (s) &speedup\\
\hline
1 &Compiled to C, no JIT &23.44 $\pm$ 0.07 &1.00\\
2 &Compiled to C, with JIT &3.58 $\pm$ 0.05 &6.54\\
3 &CPython 2.5.4 &4.96 $\pm$ 0.05 &4.73\\
\hline
\end{tabular}
\caption{Benchmarked function and results for the Python interpreter running
\texttt{f(10000000)}}
\end{figure}

@@

Classical partial evaluation has failed to be useful for dynamic language for
much the same reasons why ahead-of-time compilers cannot compile them to
efficient code. If the partial evaluator knows only the program it simply does
not have enough information to produce good code. Therefore some work has been
done to do partial evaluation at runtime. One of the earliest works on runtime
specialisation is Tempo for C \cite{consel_general_1996, consel_uniform_1996}.
However, it is essentially a normal
partial evaluator ``packaged as a library''; decisions about what can be
specialised and how, are pre-determined. Another work in this direction is DyC
\cite{grant_dyc:expressive_2000}, another runtime specializer for C. Both of these projects
have a problem similar to that of DynamoRIO.  Targeting the C language makes
higher-level specialisation difficult.

@@

A lot of work remains. We are working on two main optimizations at the moment.
Those are:

\textbf{Allocation Removal:} A key optimization for making the approach
produce good code for more complex dynamic language is to perform escape
analysis on the loop operation after tracing has been performed. In this way all
objects that are allocated during the loop and do not actually escape the loop do
not need to be allocated on the heap at all but can be exploded into their
respective fields.  This is very helpful for dynamic languages where primitive
types are often boxed, as the repeated allocation of intermediate results is
very costly.

Furthermore both tracing and leaving machine code are very slow due to a
double interpretation overhead and we might need techniques for improving those.
Eventually we will need to apply the JIT to the various interpreters that are
written in RPython to evaluate how widely applicable the described techniques
are. Possible targets for such an evaluation would be the SPy-VM, a Smalltalk
implementation \cite{bolz_back_2008}; a Prolog interpreter or PyGirl, a Gameboy
emulator \cite{camillo_bruni_pygirl:_2009}; but also not immediately obvious
ones, like Python's regular expression engine.

If these experiments are successful we hope that we can reach a point where it
becomes unnecessary to write a language specific JIT compiler and instead
possible to just apply a couple of hints to the interpreter to get reasonably
good performance with relatively little effort.
